\section{Experimental Setup}
\label{sec:experimental-setup}

\subsection{A note on the dataset behind this report}
\label{sec:dataset-note}

The dataset this section and Section~\ref{sec:results} analyze, committed at \texttt{results/sample\_run/}, is a real measurement. All three algorithm variants (the custom ring, the vendor's default \texttt{MPI\_Allreduce} selection, and \texttt{MPI\_Allreduce} forced to Open~MPI's own ring algorithm) were run against a working Open~MPI installation on the workstation described below, across the full process-count range $N = 2 \ldots 16$ and the full 8~B to 128~MiB message-size sweep. Nothing in this report is modeled, extrapolated, or filled in from a textbook constant.

Correctness was established before any timing was trusted, which matters because a benchmark of a subtly wrong allreduce produces numbers that look entirely plausible and mean nothing. The full \texttt{ctest} matrix (Section~\ref{sec:implementation}), including the non-power-of-two process counts and the zero-size-chunk cases that a smaller matrix would miss, passes against this same real Open~MPI, and every configuration's ring output is validated against both vendor \texttt{MPI\_Allreduce} and an independently computed reference before its time is recorded.

One scoping caveat belongs here rather than in a footnote: this is a single-node measurement. All ranks share one machine, so the transport underneath the point-to-point calls is Open~MPI's shared-memory path rather than a network fabric, and the fitted latency and bandwidth constants of Section~\ref{sec:cost-model} describe that path specifically. The most visible consequence, discussed in Section~\ref{sec:results}, is that effective per-byte cost is not the single constant the Hockney model assumes: message sizes whose working set stays inside the processor's 33~MB L3 cache move bytes considerably faster than sizes large enough to be bound by main-memory bandwidth. A multi-node sweep over a real fabric (\texttt{scripts/run\_slurm\_sweep.sbatch}, which the analysis pipeline consumes without modification) would be expected to follow the linear cost model far more closely, and is the single most valuable extension to this work.

\subsection{Target hardware and software}

The benchmarking machine is a workstation with an Intel Core i7-14700K (8 performance cores plus 12 efficiency cores, 28 hardware threads, and a 33~MB L3 cache that turns out to matter for the results below), 32~GB of DDR5 memory, and, unused by this CPU-only, MPI-only project, an NVIDIA RTX~5070. Because the largest process count swept is $N = 16$ and the machine exposes 28 hardware threads, no run is oversubscribed: every rank gets its own thread, which keeps the core-contention component of Section~\ref{sec:discussion}'s rank-synchronization discussion from dominating the numbers.

The MPI implementation is \textbf{Open~MPI 5.0.10} \citep{gabriel2004openmpi}, running on Ubuntu under WSL2, with GCC~15.2 and CMake. The project builds out-of-source under \texttt{-Wall -Wextra -Wpedantic} with warnings treated as errors.

\subsection{Sweep ranges and methodology}

The message-size sweep covers powers of two in bytes from $8 = 2^3$ through $134217728 = 2^{27}$ (128~MiB in the power-of-two sense \texttt{nccl-tests} uses \citep{nvidia2024ncclperf}), 25 sizes. The process-count sweep covers every integer from 2 to 16 inclusive, not only the powers of two: Section~\ref{sec:step-count}'s step-count argument applies identically to non-power-of-two $N$, and a real result there is more informative than a checkbox. Because $N$ is fixed for the lifetime of an MPI job, \texttt{scripts/run\_local\_sweep.sh} launches one \texttt{mpirun} per $N$ per MCA configuration; \texttt{apps/benchmark} itself sweeps every message size and every requested algorithm inside that one launch, so MPI startup and teardown overhead is paid once per launch rather than once per (size, algorithm) combination.

Each configuration is preceded by an \texttt{MPI\_Barrier} and a fixed number of untimed warm-up iterations (so first-touch page faults and any connection setup are not measured), then times a variable number of repetitions capped by a wall-clock budget rather than a fixed count, so small, fast configurations can accumulate many repetitions while the largest, slowest ones do not make the sweep take unreasonably long. Each repetition's elapsed time is measured with \texttt{MPI\_Wtime} on every rank and reduced to rank~0 with \texttt{MPI\_Reduce(..., MPI\_MAX, ...)}, since an allreduce is only as fast as its slowest participating rank, not as fast as whichever rank happens to be timing it. Min, mean, median, max, and standard deviation are all recorded per configuration; this report and its analysis pipeline use the \textbf{median} as "achieved" performance throughout, since it is far less sensitive than the minimum to the occasional favorable scheduling accident, while remaining far less sensitive than the mean to the occasional unfavorable one.

\subsection{Isolating implementation quality from algorithm choice}
\label{sec:mca-forcing}

\texttt{MPI\_Allreduce} is benchmarked two ways, both against the identical vendor call, and the distinction is what makes Section~\ref{sec:discussion}'s central argument checkable rather than merely plausible. The first way is Open~MPI's default \texttt{coll/tuned} algorithm selection, the real-world baseline an application actually gets, in which Open~MPI picks an algorithm per message size and process count. The second forces that selection to Open~MPI's \emph{own ring} algorithm with
\begin{center}
\texttt{--mca coll\_tuned\_use\_dynamic\_rules 1 --mca coll\_tuned\_allreduce\_algorithm 4}
\end{center}
on the \texttt{mpirun} invocation itself, giving a ring-versus-ring measurement in which both sides run the same algorithm, so any remaining difference is implementation quality rather than algorithm choice.

The numeric algorithm id was read from the installed library rather than assumed, since it has moved across Open~MPI releases. On Open~MPI 5.0.10, \texttt{scripts/check\_mpi\_algorithms.sh} reports the \texttt{coll\_tuned\_allreduce\_algorithm} enumerator as \texttt{0:ignore}, \texttt{1:basic\_linear}, \texttt{2:nonoverlapping}, \texttt{3:recursive\_doubling}, \texttt{4:ring}, \texttt{5:segmented\_ring}, \texttt{6:rabenseifner}, so \textbf{4} is the ring. That the enumerator also exposes \texttt{recursive\_doubling} and \texttt{rabenseifner} as separate selectable algorithms is itself worth noting: those are precisely the $O(\log N)$-step algorithms Section~\ref{sec:step-count} argues a tuned collective will prefer over a ring at small message sizes, and the default-selection variant is free to choose them.

Because MCA parameters are process-environment state read once at \texttt{MPI\_Init}, forcing the ring algorithm is a property of how \texttt{mpirun} was invoked for a given launch, not something \texttt{apps/benchmark} can flip mid-run; \texttt{scripts/run\_local\_sweep.sh} therefore performs two \texttt{mpirun} invocations per $N$, one plain and one with the forcing flags, appending both to the same results file.
